Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-07-25 paper.tex
Updated methods sub-section on cluster reads

% new commands
\newcommand{\todo}[1]{
  \color{red}\emph{[#1]}
  \color{black}
}

@@

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (x,y,z)
representation. That is, x slices are written one after the other, and
in every x slice, y rows are written one after the other. NifTI, the
file format defined by the Neuroimaging Informatics Technology
Initiative Nifti\footnote{\url{https://nifti.nimh.nih.gov}} is such a
format, used by the vast majority of the neuroimaging community. In
the remainder, we use the term ``slice" for ``x slice".

@@

The main idea of cluster reads is to load multiple blocks in memory,
concatenate them in a buffer and write the buffer in the reconstructed
image. Seeking is reduced compared to naive block merging since
contiguous parts of the buffer will be written without seeking. In
cluster reads, a given block is accessed only once. That is, the
amount of used memory $m'$ is such that $m'=kd^3b, k \in \mathbb{N}$.

The buffer might be a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up when two blocks of one slice
and three rows per slice are loaded. Implementation details about the
buffer are given later.
\begin{figure}
\centering
\def\svgwidth{\columnwidth}
\input{figures/buffer-cluster-reads.pdf_tex}
\caption{Illustration of the buffer used in cluster reads. White
  portions are not allocated. [a,b] refers to block a, row b.}
\label{fig:cluster-reads-buffer}
\end{figure}

@@

\todo{Check that ordering in the Figure is consistent with Nifti's.}

Incomplete block rows and incomplete block slices lead to extra
seeking and should be avoided as much as possible.  Thus, cluster
reads consider only three cases, represented in
Figure~\ref{fig:cluster-reads}:
\begin{itemize}
  \item Case 1: $m<d^3\sqrt[3]{n}b=\frac{D^3}{\sqrt[3]{n}^2b}$, that is, a
    row of blocks does not fit in memory. Incomplete rows of
    blocks will be merged, with $d^2$ seeks per memory load.
  \item Case 2: $\frac{D^3}{\sqrt[3]{n}^2}b\leq m<\frac{D^3}{\sqrt[3]{n}}b$,
    that is, a row of blocks fits in memory but a slice of blocks does
    not. Only complete rows of blocks will be merged, with
    $d$ seeks per memory load. We have $m'=\floor*{\frac{m}{d^3\sqrt[3]{n}b}}$.
    \item Case 3: $\frac{D^3}{\sqrt[3]{n}}b \leq m < D^3b$, that is, a slice
      of blocks fits in memory. Only complete slices of blocks
      will be merged, a single seek per memory load. We have
      $m'=\floor*{\frac{m}{d^3\sqrt[3]{n}^2b}}$.
\end{itemize}

Algorithm~\ref{algo:cluster-reads} shows the resulting cluster reads
method. Function \texttt{adjust\_memory} (line 4) returns $m'$
according to the three cases above. Function \texttt{sizeof} (line 6)
returns the memory used by the buffer, including only its allocated
segments. It should be noted that writing the buffer to the
reconstructed image (line 11) might involve some manipulation in the
buffer to merge its continuous segments. 
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \STATE m'=adjust\_memory(m)    
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m'}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3.pdf_tex}

\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
% See source files in
% https://drive.google.com/drive/folders/0BzT7pbvPDUdQRXppVlhfdS0ycGc?usp=sharing
\end{figure}

The total number of seeks performed by cluster reads depends on the three
cases listed above and on the overall number of memory loads.
\begin{equation}
  N_\mathrm{cluster\_reads\_v1}= n + \frac{bR}{m'}d^2+d^2\alpha
\end{equation}
Similarly, in Case 2 we have:
\begin{equation}
  N_\mathrm{cluster\_reads\_v2}= n + \frac{bR}{m'}d+d\alpha
\end{equation}
And in Case 3:
\begin{equation}
  N_\mathrm{cluster\_reads\_v3}= n + \frac{bR}{m'}
\end{equation}

@@

In multiple reads, blocks are read partially to ensure that the memory
buffer only contains contiguous bytes, at any time. That is, a given
block will be read multiple times.

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads
    \todo{improve notations and algo}.}
  \label{algo:multiple-reads}
  \begin{algorithmic}
  \STATE sorted\_split\_name\_list = sort blocks by increasing (x,y,z) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

>>>>

% new commands
\newcommand{\todo}[1]{\marginpar{\parbox{18mm}{\flushleft\tiny\color{red}\textbf{TODO}:
      #1}}}

@@

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (x,y,z)
representation. That is, x slices are written one after the other, and
in every x slice, y rows are written one after the other. NIfTI, the
file format defined by the Neuroimaging Informatics Technology
Initiative NIfTI\footnote{\url{https://nifti.nimh.nih.gov}} is such a
format, used by the vast majority of the neuroimaging community. In
the remainder, we use the term ``slice" for ``x slice".

@@

The main idea of cluster reads is to load multiple blocks in memory,
concatenate them in a buffer and write the buffer in the reconstructed
image. Seeking is reduced compared to naive block merging since
contiguous parts of the buffer will be written without seeking. In
cluster reads, a given block is accessed only once.

The buffer might be a
slightly complex data structure, for instance an associative array or
a Python dictionary, capable of storing multiple disjoint sequences of
contiguous bytes without having to allocate memory for the bytes
between such sequences. Figure~\ref{fig:cluster-reads-buffer}
illustrates how the buffer would fill up for the two first blocks in a reconstructed image.Implementation details about the
buffer are given later.
\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/buffer.pdf_tex}
\caption{Illustration of the buffer used in cluster reads, showing how
  the two first 4x4 blocks of a reconstructed image would be arranged
  in the buffer.  White portions in the buffer are not allocated.  The
  column-major order is used, similar to the NIfTI format: i is the
  fastest changing dimension, followed by j, followed by k.}
\label{fig:cluster-reads-buffer}
\end{figure}

@@

\begin{figure}
\centering
\def\svgwidth{0.3\columnwidth}
\input{figures/case1-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case2-a.pdf_tex}
\def\svgwidth{0.3\columnwidth}
\input{figures/case3.pdf_tex}
\caption{Memory-load configurations in cluster reads, leading to
  different number of seeks. Red blocks need seeking before each of
  their rows ($d^2$ seeks). Blue blocks need seeking before each of
  their slices ($d$ seeks in total). Green blocks need only a single
  seek. Grey, dashed, transparent blocks represent the contiguous
  memory loads and are added for the sake of visualization.}
\label{fig:cluster-reads}
% See source files in
% https://drive.google.com/drive/folders/0BzT7pbvPDUdQRXppVlhfdS0ycGc?usp=sharing
\end{figure}
Our cluster reads algorithm focuses on the three memory load
configurations represented in Figure~\ref{fig:cluster-reads}, that is,
the amount of memory $m'$ used by the algorithm is rounded down to the
closest number of complete blocks (case 1), complete block rows (case
2) or complete block slices (case 3). This is in general reasonable
since adding an incomplete row to a set of complete ones multiplies
the number of required seeks by $d$, as illustrated in
Figure~\ref{fig:avoided-configurations}-Left. In some cases though,
rounding $m$ down to $m'$ might increase the number of required
memory load to a point that the overall number of seeks also
increases. However, such cases are a bit pathological and their
complete description requires extensive calculations involving modulo
arithmetic, which we felt were unwieldy to report here.
\begin{figure}[b]
  \centering
\def\svgwidth{0.3\columnwidth}
\input{figures/incomplete-rows.pdf_tex}
\hfill
\def\svgwidth{0.3\columnwidth}
\input{figures/overlap.pdf_tex}
\caption{Configurations that increase the number of seeks per
  memory-load and are thus deliberately avoided by cluster
  reads. Left: configuration with incomplete block rows (multiplies
  number of seeks by $d$). Right: configuration with block rows that
  overlap multiple block slices (multiplies number of seeks by $2$).}
\label{fig:avoided-configurations}
\end{figure}
To summarize, the amount of memory used $m'$ is set as follows in each of the 3 cases:
%% \begin{equation*}
%%   m_{case 1}' = \floor*{\frac{m}{d^3b}}; m_{case 2}' = \floor*{\frac{m}{d^3b\sqrt[3]{n}}}; m_{case 3}' = \floor*{\frac{m}{d^3b\sqrt[3]{n}^2}}
%% \end{equation*}
%% which gives:
\begin{equation*}
  m_1' = \floor*{\frac{mn}{Rb}};
  m_2' = \floor*{\frac{m\sqrt[3]{n}^2}{Rb}};
  m_3' = \floor*{\frac{m\sqrt[3]{n}}{Rb}}
\end{equation*}

In addition, our algorithm avoids configurations where the memory load
overlaps multiple block slices in case 2 or multiple block rows in
case 1, as such overlaps multiply the number of required seeks
(see Figure~\ref{fig:avoided-configurations}-Right). Thus, the total
number of memory loads $x$ required to reconstruct the image is:
\begin{equation*}
  x_1 = \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{n}^2;
  x_2 = \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{n};
  x_3 = \ceil*{\frac{Rb}{m_3'}}
\end{equation*}

Finally, the total number of seeks performed by cluster reads to reconstruct the image is:
%% \begin{equation*}
%% N_\mathrm{cr} =
%% \begin{cases}
%%   n + \sqrt[3]{n}^2 \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}d^2 & \text{if } \frac{m}{d^3b} < \sqrt[3]{n}\\
%%   n + \sqrt[3]{n} \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}d      & \text{if } \sqrt[3]{n} \leq \frac{m}{d^3b} < \sqrt[3]{n}^2\\
%%   n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{m}{d^3b} < n\\
%%   n + 1                                                          & \text{otherwise}
%% \end{cases} \label{eq:seeks-cluster-reads}
%% \end{equation*}
%% And since $d=\frac{D}{\sqrt[3]{n}}$ and $R=D^3$, we finally have:
\begin{equation}
N_\mathrm{cr} =
\begin{cases}
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}^2m_1'}}\sqrt[3]{R}^2     & \text{if } \frac{mn}{Rb} < \sqrt[3]{n}\\
  n + \ceil*{\frac{Rb}{\sqrt[3]{n}m_2'}}\sqrt[3]{R}        & \text{if } \sqrt[3]{n} \leq \frac{mn}{Rb} < \sqrt[3]{n}^2\\
  n + \ceil*{\frac{Rb}{m_3'}}                              & \text{if } \sqrt[3]{n}^2 \leq \frac{mn}{Rb} < n\\
  n + 1                                                          & \text{otherwise}
\end{cases} \label{eq:seeks-cluster-reads-1}
\end{equation}

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 4)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
true.  Function \texttt{sizeof} (line 8) returns the actual memory
used by the buffer, including only its allocated segments. It should
be noted that writing the buffer to the reconstructed image (line 11)
might involve some manipulation in the buffer to merge its continuous
segments. \todo{Precise this based on Valerie's final code.}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE list = sort blocks by increasing (z,y,x) values
    \STATE initialize buffer
    \STATE (m',case)=\texttt{switch}(m,n,R,b)
    \STATE old\_block = list[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = list[i]
      \STATE overlap = \texttt{check\_overlap}(block,old\_block,case)
      \IF{\texttt{sizeof}(buffer) \textless m' \textbf{and} overlap=false}
      \STATE read block and insert it in buffer
      \ELSE
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

@@

In multiple reads, blocks are read partially to ensure that the memory
buffer only contains contiguous bytes, at any time. That is, a given
block will be read multiple times.     \todo{improve notations and algo in algorithm}

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads.}
  \label{algo:multiple-reads}
  \begin{algorithmic}
  \STATE sorted\_split\_name\_list = sort blocks by increasing (x,y,z) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE
